--- 
title: "Example analysis of a phyloseq object with microbiomeX"
author: "CBMR Team"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
classoption: a4paper
#bibliography: [book.bib, packages.bib]
#biblio-style: apalike
link-citations: yes
# description: "MicrobDiab Analysis"
#output: bookdown::pdf_book
#  fig_width: 7
---




# Introduction

The document uses example data to illustrate how analyses of a phyloseq object can be done with microbiomeX.

How it works:

- you need a phyloseq object saved as **.rds** object (via saveRDS).
    - the sample_data of your phyloseq object needs to contain a variable to group the samples (the name of that column will be your group_var) 
- In the section **Load phyloseq object and set your input parameters for the analysis** you define the path to your phyloseq object and set important input parameters such as group_var. 
- in the end you get a nice html file



# Load packages, and two color schemes

<!-- Load required packages -->

```{r, echo = FALSE, message=FALSE, include = FALSE}
# - install packages -
# NB: all packages have to be installed, usually via install.packages()
# but some you need to install via bioconductor, e.g.:
# source("https://bioconductor.org/biocLite.R")
# biocLite("phyloseq")
# --

# - load packages -
library(phyloseq); # packageVersion("phyloseq")
library(dada2); # packageVersion("dada2")
library(vegan); # packageVersion("vegan")
library(ggplot2); # packageVersion("ggplot2")
library(dplyr); # packageVersion("dplyr")
library(tidyr); # packageVersion("tidyr")
library(gridExtra); # packageVersion("gridExtra")
library(xtable); # packageVersion("xtable")
library(RVAideMemoire); #packageVersion("RVAideMemoire")
library(viridis); # packageVersion("viridis")
library(scales); # packageVersion("scales") # for the oob = squish option in gradient plots
library(ggthemes); # packageVersion("ggthemes")
library(DESeq2); # packageVersion("DESeq2")
library(ggpubr); # packageVersion("ggpubr")

# - packaged from Mani's analysis that I could add -
library(cowplot)
library(coin)
library(ade4)
library(ggrepel)
library(ggforce)
library(bookdown)
library(kableExtra)
library(RColorBrewer)
library(bindrcpp)
library(stringr)
library(pheatmap)
library(BiocParallel)
library(microbiomeX)
# --

# - define cbPalette and QuantColors15 schemes -
# R color blind palette is used for group comparisons (unless more than 8 groups)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7") # ref: http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/
# QuantColors15 is used for phyla comparisons as long as there are < 15 phyla, each color is easy to distinguish from the two colors surrounding it
tol15rainbow=c("#114477", "#4477AA", "#77AADD", "#117755", "#44AA88", "#99CCBB", "#777711", "#AAAA44", "#DDDD77", "#771111", "#AA4444", "#DD7777", "#771144", "#AA4477", "#DD77AA") # ref: https://tradeblotter.wordpress.com/2013/02/28/the-paul-tol-21-color-salute/
QuantColors15 <- tol15rainbow[c(1, 12, 4, 7, 13, 2, 11, 5, 8, 14, 3, 10, 6, 9, 15)]
# --

# - define style options for the html document -
options(knitr.table.format = "latex")
# --
```



# Load your phyloseq object and set and test your input parameters

- all lines where user user input is required or possible are marked with NB: USER INPUT and NB: user input, respectively


```{r, echo = FALSE, message=FALSE}
# - source all functions -
functionpath <- "./Functions"
source(file.path(functionpath, "000_helper_functions.R"))
source(file.path(functionpath, "010_explore_ps_functions.R"))
source(file.path(functionpath, "020_preprocess_filtering_functions.R"))
source(file.path(functionpath, "030_alpha_diversity_functions.R"))
source(file.path(functionpath, "040_beta_diversity_functions.R"))
source(file.path(functionpath, "050_diff_abundance_functions.R"))
source(file.path(functionpath, "060_phylum_analysis_functions.R"))

# - load your phyloseq object - 
# NB: USER INPUT
name_phyloseq_rds <- "microdiab_DK.rds"
# NB: USER INPUT
datapath <- "./Example_phyloseq_objects"

# Test your input:
if(! file.exists(file.path(datapath, name_phyloseq_rds))){
        stop("The given path/file to your phyloseq object does not exist.")
}

ps <- readRDS(file.path(datapath, name_phyloseq_rds))

# --

# - Define the group variable for sample comparisons -
# NB: USER INPUT
group_var <- "Status" # MANDATORY: a variable in sample_data(ps) based on which samples will be compared

# Test your input:
if(! group_var %in% colnames(sample_data(ps))) {
        stop("The given group_var is not a variable in the sample data of the loaded phyloseq object.")
}
# --

# - Define the order of the groups in your group_var -
# NB: USER INPUT
group_var_levels <- c("NGT", "T2D") # defines the order of the groups in all plots. If set to NULL:
if (is.null(group_var_levels)) { 
        group_var_levels <- unique(sample_data(ps)[[group_var]])
}

# Test your input:
if (!all(group_var_levels %in% unique(sample_data(ps)[[group_var]])) ||
    !all(unique(sample_data(ps)[[group_var]]) %in% group_var_levels)) {
        stop("The given group_var_levels do not fit to all unique values in the group_var column.")
}

if (length(group_var_levels) < 2) {
        stop("Sorry, you currently need to have at least two levels in your group_var for this analysis to make sense.")
}

if (length(group_var_levels) > 15) {
        stop("Sorry, there are currently max 15 levels (groups) allowed in your group_var.")
}
# -- 

# - Set the colors representing the groups in your group_var -
# NB: USER INPUT
color_levels <- c(cbPalette[2], cbPalette[4]) # choose your preferred colors for each group in your group_var. If set to NULL:
if (is.null(color_levels)) { 
        if (length(group_var_levels) < 9) {
                color_levels <- cbPalette[1:length(group_var_levels)]
        } else {
                color_levels <- QuantColors15[1:length(group_var_levels)]
        }
}

# Test your input:
if (length(color_levels) != length(group_var_levels)) {
        stop("The number of colors in the given color_levels did not fit to group_var_levels.")
}

if (!all(areColors(color_levels))) {
        stop("Stopped because not all entries in color_levels were valid R colors.")
}
# -- 

# - optional: use another variable in sample_data(ps) as shape in some of the plots -
# NB: user input
shape <- NULL 

# Test your input:
if (!is.null(shape)) {
        if (! shapoe %in% colnames(sample_data(ps))) {
                stop("The given shape is not a variable in the sample data of the loaded phyloseq object.")
        }
}
# --

# - Set the taxonomic_level for the analysis -
# NB: You can choose all levels within colnames(tax_table(ps))
# phyloseq::taxglom to that level will be used.
# if you set taxonomic_leve to NULL, all taxa will be analysed (i.e. strain/species/ASV level analysis)
# NB user input
taxonomic_level <- NULL
# NB user input
phylum_NA_removal <- "no" # if yes you remove all taxa for which Phylum = NA before the analyses

# Test your input:
if (!is.null(taxonomic_level) && !(taxonomic_level %in% colnames(tax_table(ps)))){
        stop("You provided a taxonomic_level that is not present in the tax_table of the loaded phyloseq object.")
}
# --

# - Filtering inputs -
# NB: user input
prevalence <- 5 # in percent, for prevalence filtering of taxa 
# NB: user input
min_obs <- 0L # in counts, a taxon will be considered present (for prevalence) if count > min_obs
# NB: user input
taxa_sums_quantile <- 90 # in percent, taxa whose taxa_sums are above this threshold will be kept even if they do not pass prevalence filter

# Test filtering inputs and set invalid inputs to defaults
if (!is.numeric(prevalence) || prevalence < 0 || prevalence > 100) {
        prevalence <- 5
}

if (!is.numeric(min_obs)) {
       min_obs <- 0L
}

if (!is.numeric(taxa_sums_quantile) || taxa_sums_quantile < 0 || taxa_sums_quantile > 100) {
        taxa_sums_quantile <- 0
}
# --

# - Alpha diversity and rarifaction inputs -
# NB: user input
alpha_div_measures <- c("Observed", "Shannon") # ?phyloseq::estimate_richness: may include c("Observed", "Chao1", "ACE", "Shannon", "Simpson", "InvSimpson", "Fisher")
# NB: user input
seed <- 1234 # for reproducibility of rarefaction curves
# NB: user input
step_size <- 200 # steps in sample_sums for rarefaction curves
# NB: user input
rare_level <- NULL # sample_sums/total_counts level to which rarefaction is perforemd, if NULL, min(sample_sums(ps)) is used!
# NB: user input
rare_max_total <- NULL # maximal sample_sums/total_counts value to which rarefaction curves are calculated, if NULL: quantile(sample_sums(ps), probs = .25) is used

# Test alpha-diversity inputs and set invalid inputs to defaults
if (! all(alpha_div_measures %in% c("Observed", "Chao1", "ACE", "Shannon", "Simpson", "InvSimpson", "Fisher"))) {
        alpha_div_measures <- c("Observed", "Shannon")
}

if (!is.numeric(seed)) {
        seed <- 1234
}

if (!is.numeric(step_size)) {
        step_size <- 200
}

if (!is.numeric(rare_level)) {
        rare_level <- NULL # will be changed to min(sample_sums(ps))
}

if (!is.numeric(rare_max_total)) {
        rare_max_total <- NULL # quantile(sample_sums(ps), probs = .25)
}
# --

# - beta diversity inputs -
dist_methods <- c("jsd") # otions: see unlist(phyloseq::distanceMethodList)
coord_cor <- FALSE # adjustment of the axis representation in ordination plots

# Test beta diversity inputs and set invalid inputs to defaults
if (! all(dist_methods %in% unlist(phyloseq::distanceMethodList))) {
       dist_methods <- c("jsd")
}

if (! is.logical(coord_cor)){
        coord_cor <- FALSE
}
# --
```


<!-- Command for building the book in the end -->

```{r, eval = FALSE, include=FALSE}
# Use this command for building the book in the end.
bookdown::render_book("index.Rmd", "bookdown::pdf_book")
#rmarkdown::render_site(output_format = 'bookdown::pdf_book', encoding = 'UTF-8')
```



# Print package versions and input parameters


```{r, echo = FALSE, message=FALSE}
# - record used R and package versions in a data frame -
PackageVersions <- data.frame(R = version$version.string,
                              phyloseq = packageVersion("phyloseq"),
                              dada2 = packageVersion("dada2"),
                              vegan = packageVersion("vegan"),
                              ggplot2 = packageVersion("ggplot2"),
                              dplyr = packageVersion("dplyr"),
                              tidyr = packageVersion("tidyr"),
                              gridExtra = packageVersion("gridExtra"),
                              xtable = packageVersion("xtable"),
                              viridis = packageVersion("viridis"),
                              scales = packageVersion("scales"),
                              ggthemes = packageVersion("ggthemes"),
                              DESeq2 = packageVersion("DESeq2"),
                              ggpubr = packageVersion("ggpubr"))

for (i in 1:ncol(PackageVersions)){
        PackageVersions[,i] <- as.character(PackageVersions[,i])
}

PackageVersions_l <- tidyr::gather(PackageVersions, key = Package, value = Version) 

PackageVersions_l[2:nrow(PackageVersions_l), ] <- dplyr::arrange(PackageVersions_l[2:nrow(PackageVersions_l), ], Package)
# --

# - record given input parameters in a data frame -
Inputs_phyloseqAnalysis <- list(taxonomic_level = taxonomic_level,
                                phylum_NA_removal = phylum_NA_removal,
                                group_var = group_var,
                                group_var_levels = group_var_levels,
                                shape = shape,
                                prevalence = prevalence,
                                min_obs = min_obs,
                                taxa_sums_quantile = taxa_sums_quantile,
                                alpha_div_measures = alpha_div_measures,
                                seed = seed,
                                step_size = step_size,
                                rare_level = rare_level,
                                rare_max_total = rare_max_total,
                                dist_methods = dist_methods,
                                coord_cor = coord_cor)


Inputs_phyloseqAnalysis <- data.frame(Argument = names(Inputs_phyloseqAnalysis),
                                      Value = sapply(Inputs_phyloseqAnalysis, function(x){
                                              if (is.null(x)){
                                                      "NULL"
                                              } else if (length(x) > 7) {
                                                      "long entry"
                                              } else if (length(x) > 1 && length(x) < 8){
                                                      paste(x, collapse = "; ")
                                              } else {
                                                      as.character(x)
                                              }
                                      }
                                      ))

rownames(Inputs_phyloseqAnalysis) <- NULL
# --
```


- the following package veersions have been used:

```{r, echo = FALSE, message = FALSE}
knitr::kable(PackageVersions_l)
```


- the following inputs were given

```{r, echo = FALSE, message = FALSE}
knitr::kable(Inputs_phyloseqAnalysis)
```




