---
title: "McMurdieLikeSimulation"
author: "Thorsten Brach"
date: "11/12/2017"
output: html_document
---

# Packages

```{r, message = FALSE, warning =FALSE, include=FALSE, cache=FALSE}
library(phyloseq); packageVersion("phyloseq")
library(DESeq2); packageVersion("DESeq2")
library(knitr); packageVersion("knitr")
library(ggplot2); packageVersion("ggplot2")
library(dplyr); packageVersion("dplyr")
library(tidyr); packageVersion("tidyr")
library(lubridate); packageVersion("lubridate")
library(gridExtra); packageVersion("gridExtra")
library(viridis); packageVersion("viridis") # for gradient plots
library(scales); packageVersion("scales") # for the oob = squish option in gradient plots
library(ggthemes); packageVersion("ggthemes") # 

```



# Source Functions and load physeq object

```{r, message = FALSE, warning =FALSE}
datapath <- "/Users/jvb740/MarieCurie_Work/BackgroundKnowledge/16S_Learning/Dada_Pipel/TbT_SimulationData"

physeq <- readRDS(file.path(datapath, "Mouse_YOMA_physeq.rds"))

# source(file.path(datapath, "170514_SimulationFunctions.R"))
source(file.path(datapath, "TbT_Simulation_Functions.R"))

# change the taxa names to something simpler
taxa_names(physeq) <- paste0("Taxon_", 1:ntaxa(physeq))

# savepath <- "/Users/jvb740/MarieCurie_Work/relativeAbundanceGame/RASimulation/Simulations"
```

- the **unfiltered** input physeq contains **`r ntaxa(physeq)` taxa and `r nsamples(physeq)` samples**
- The **sparsity in percentage is `r round(100*(sum(otu_table(physeq) == 0)/(ntaxa(physeq)*nsamples(physeq))), 3)`**



# Taxa filtering on the original physeq


##  Setting the filtering parameters

```{r, message = FALSE, warning =FALSE}
sampletypes <- "Mouse-YOMA" # in most cases you will only have one sample type

prevalence <- 20

# an option to use a different prevalence filter for the simulated data
prevalence_sim <- 10 # if NULL prevalence will be used

# Minimum number of reads above which to consider a taxon "observed" in a sample, for prevalence calculation
minobs <- 0L # changed to 0 so that 1 count is already present

# MinTaxaSum: taxa stay in the physeq even when not fulfilling the prevalence criterion if their taxa_sum is abouve MinTaxaSum
MinTaxaSum <- quantile(taxa_sums(physeq), probs = 0.97) # MinTaxaSum is the 90% quantile of the taxa sums, so
# you would not kick out any taxon that is in the top 10% of taxa_sums
# if (MinTaxaSum < 1000){MinTaxaSum <- 1000}
```


##  Filter the input physeq to become the actual simulation template

- NB: the McMurdie code allowed for several sampletypes, therefore the list here

```{r, message = FALSE, warning =FALSE}
physeq <- filter_taxa(physeq, function(x){(sum(x > minobs) > (prevalence/100)*length(x)) || (sum(x) > MinTaxaSum)}, prune = TRUE)

# equal to or check with: 
# sum(apply(as(otu_table(physeq), "matrix"), 2, function(x){(sum(x > minobs) > (prevalence/100)*length(x)) || (sum(x) > MinTaxaSum)}))

templatelist <- list(physeq)
names(templatelist) <- sampletypes
```

- the final filtered input physeq contains **`r ntaxa(physeq)` taxa and `r nsamples(physeq)` samples**
- The **sparsity in percentage is `r round(100*(sum(otu_table(physeq) == 0)/(ntaxa(physeq)*nsamples(physeq))), 3)`**



# Simulating the data


##  Setting the simulation parameters

```{r, message = FALSE, warning =FALSE}
seed <- sample(10^8, size = 1)
set.seed(seed)

# Vector of the replicate numbers for each combination of simulation parameters
reps <- 1:2 #  i.e how many simulations will be generated

# Define the number of true-positive taxa (will be the same in each simulation) 
nTP <- 9 

# Define the number of samples for each group (will be the same in each simulation)
J1 <- 25
J2 <- 25

# Define the range from wich the foldeffect (effect size) will be sampled (will be the same in each simulation)
foldeffect <- 3:6

# The average number of reads per simulated sample:
ns <- median(sample_sums(physeq)) # with ns = median(sample_sums(physeq)) the library sizes of the simulated samples will be sampled from sample_sums(physeq). 

# The delimiter in the command parameter string
comdelim = "_"

# # The number of cores to use in this simulation
# Ncores <- 30

# - generate simparams from the given parameters -
if(is.numeric(ns)){
        # Define the simulation parameters combinations
        simparams <- apply(expand.grid(ns, sampletypes, reps, foldeffect[1],
                                      foldeffect[length(foldeffect)],J1, J2), 1, paste0, collapse=comdelim)
} else {
        stop("ns must be numeric.")
}

# Define the labels to go with each element of the simulation parameter after splitting on the delimiter
simparamslabels <- c("nreads", "SampleType", "Replicate", "minEffectSize", "maxEffectSize", "nsamplesgr1", "nsamplesgr2")

datestamp1 <- now()
```


## Run simulation to get the simulated count tables

```{r, message = FALSE, warning =FALSE}
ptm <- proc.time()
simlist <- simulate_counts_withTPTaxa(templatelist = templatelist, simparams = simparams, simparamslabels = simparamslabels, nTP = nTP)
proc.time() - ptm
```




