---
title: "170106_BetaDiversityDistancesOrdination"
author: "Thorsten Brach"
date: "6/1/2017"
output: html_document
---

# Packages
```{r}
try(library(dada2), biocLite("dada2"))

## Short Read
try(library(phyloseq), biocLite("phyloseq"))

## ggplot2
try(library(ggplot2), install.packages("ggplot2"))

## dplyr
try(library(dplyr), install.packages("dplyr"))

## dplyr
try(library(tidyr), install.packages("tidyr"))

```

# Background

We want to measure and illustrate how similar the microbial communities are in different samples. Alpha-diversity measures tell us how diverse a individual sample is, now we want to know how similar/distant different samples are to each other. Are the samples from patients more similar to each other than to healthy controls?

The first step is therefore to calculate distances between the samples. 
The second step is then to illustrate the distances in a two or three dimensional coordinate frame.

# Calculating distances between microbial communities

## Preprocessing the Global Patterns example data

- this sample data contains human samples from Feces, Mock, Skin, and Tongue, and you would expect that samples from the same "area" should be more similar to each other
- NB: we change to relative abundances here on the way, because some distance measures seem to only work on relative abundances

```{r}
## example physeq
data(GlobalPatterns)
GP <- GlobalPatterns
# ntaxa(GP) # 19216

## Filtering the GP to reduce processing time
# keep only human samples
sample_data(GP)$human = factor( get_variable(GP, "SampleType") %in% c("Feces", "Mock", "Skin", "Tongue") )

GPh <- subset_samples(GP, subset = human == TRUE )

## keep only taxa present in at least 30% of samples
GPhf <- filter_taxa(GPh, function(x){sum(x != 0) > .3*length(x)}, prune = TRUE) # still 3670 taxa

## do their normalisation
total = median(sample_sums(GPhf))
standf = function(x, t=total) round(t * (x / sum(x)))
GPhfs = transform_sample_counts(GPhf, standf)
#range(sample_sums(GPhfs))

## change to relative abundance here
GPhfs = transform_sample_counts(GPhfs, function(x){x/sum(x)})

## keep only taxa with good variation
GPhfsf = filter_taxa(GPhfs, function(x) sd(x)/mean(x) > 3.0, prune = TRUE) # down to 388 taxa
sample_sums(GPhfsf) # again very different but I think fine since all have now same taxa

```

```{r}
plot_bar(GPhfsf, x = "Sample", y = "Abundance", fill = "SampleType")
plot_bar(GPhfsf, x = "SampleType", y = "Abundance", fill = "Phylum")

```

## Calculate different distances and understand at least some

```{r}
# phyloseq offers a lot of distance methods
dist_methods <- unlist(distanceMethodList)
# I reduce
dist_methods <- dist_methods[dist_methods %in% c("unifrac", "wunifrac", "jsd", "canberra", "bray", "jaccard", "euclidean")]

dist_methods
```

### Calculate the distance matrixes with phyloseq and try to recapitulate some of them

- NB: UniFrac requires a phylogenetic tree that is present in this sample data

```{r}

# NB: UniFrac requires a tree! GP has a tree

Distlist <- vector("list", length(dist_methods))
names(Distlist) = dist_methods

for( i in dist_methods ){
        # Calculate distance matrix
        iDist <- phyloseq::distance(GPhfsf, method=i)
        Distlist[[i]] = iDist
}

```

#### recapitulate Euclidean distance

```{r}
# phyloseq makes use here of the vegan package
DistEucl <- vegan::vegdist(t(as(otu_table(GPhfsf), "matrix")), method = 'euclidean')

identical(as(DistEucl, "matrix"), as(Distlist$euclidean, "matrix"))

# the function (type vegan::vegdist) is unfortunately based on C code
# but ?vegan::vegdist shows you
# euclidean	d[jk] = sqrt(sum(x[ij]-x[ik])^2) # but actually it is sqrt(sum((x[ij]-x[ik])^2))
# THINK of a2 + b2 = c2, then c <- sqrt(a2 + b2)
# j and k refer to the two samples to be compared
# i to the current taxa

##### Own Euclidean distance
# First example for only a few samples
physeq <- GPhfsf
if(taxa_are_rows(physeq)){
        DF <- as.data.frame(otu_table(physeq))     
} else {
        DF <- as.data.frame(t(otu_table(physeq)))
}

colnames(DF)[1:2] # "M31Fcsw" "M11Fcsw"
Dist12 <- sqrt(sum((DF[,1]-DF[,2])^2))
all.equal(Dist12, DistEucl[1]) # TRUE!


# how to implement this, basically column wise calculation
physeq <- GPhfsf
if(taxa_are_rows(physeq)){
        DF <- as(otu_table(physeq), "matrix")     
} else {
        DF <- as(t(otu_table(physeq)), "matrix")
}

## Long way (computationally) with apply and sapply:
EuclDistMatrix <- sapply(1:ncol(DF), function(y){apply(DF, 2, function(x){sqrt(sum((x-DF[,y])^2))})})
colnames(EuclDistMatrix) <- rownames(EuclDistMatrix)
# compare: 
DistEuclM <- as(DistEucl, "matrix")
all.equal(DistEuclM, EuclDistMatrix) # TRUE
```

- NB: I like the apply way here, but it is of course not computationally super efficient, Check:
- <http://stackoverflow.com/questions/34757217/fastest-way-to-apply-function-to-all-pairwise-combinations-of-columns>
- thus the data.table package might be smart to learn (in c), I could not figure it out easily here, at for small sample sizes I guess the apply solution is good enough

```{r}
# library(data.table)
# data.DT <- setkey(data.table(c(DF), colId = rep(1:ncol(DF), each = nrow(DF)), rowId = rep(1:nrow(DF), times = ncol(DF))), colId)
# 
# diff.DT <- data.DT[
#   , {
#     ccl <- unique(colId)
#     vv <- V1
#     data.DT[colId > ccl, .(col2 = colId, V1 - vv)]
#   }
#   , keyby = .(col1 = colId)
# ]
```


#### recapitulate Bray Curtis distance

```{r}
# phyloseq makes also here use of the vegan package
DistBray <- vegan::vegdist(t(as(otu_table(GPhfsf), "matrix")), method = 'bray')

identical(as(DistBray, "matrix"), as(Distlist$bray, "matrix"))

##### Own Bray distance distance
# from ?vegan::vegdist
# bray	d[jk] = (sum abs(x[ij]-x[ik]))/(sum (x[ij]+x[ik]))

physeq <- GPhfsf
if(taxa_are_rows(physeq)){
        DF <- as(otu_table(physeq), "matrix")     
} else {
        DF <- as(t(otu_table(physeq)), "matrix")
}

## Long way (computationally) with apply and sapply:
BrayDistMatrix <- sapply(1:ncol(DF), function(y){apply(DF, 2, function(x){ sum(abs(x-DF[,y]))/sum(x + DF[,y]) })})
colnames(BrayDistMatrix) <- rownames(BrayDistMatrix)
# compare: 
DistBrayM <- as(DistBray, "matrix")
all.equal(DistBrayM, BrayDistMatrix) # TRUE
```

### Summary Distances

- you should really think about which distance could make most sense for your data and understand the ideas behind some of them
- at least the implementation should be pretty easy
- so know you have distances between the different samples
    - ** Are these all beta diversity measures?**
- How to use those distances to cluster the samples in some dimensions?

# Ordination: Using the distances to cluster the samples in two dimensions
