---
title: "170120_EffectOfRefDatabase"
author: "Thorsten Brach"
date: "20/1/2017"
output: html_document
---

# Required packages

```{r, message = FALSE}
library(phyloseq); packageVersion("phyloseq")
library(DESeq2); packageVersion("DESeq2")
library(knitr); packageVersion("knitr")
library(ggplot2); packageVersion("ggplot2")
library(dplyr); packageVersion("dplyr")
library(gridExtra); packageVersion("gridExtra")

```

# Background

- checking how the taxanomic assignment of abundant taxa in the 243 DanFunD sample seqtab depends on the reference databases given on the Dada2 tutorial website
- the seqtab had been filtered (to f_seqtab) so that taxa remained that were present in at least 20% of the samples
    - see below: min(colSums(seqtab > 0)/nrow(seqtab))


```{r}
# # Dada_WrapperAssignTaxonomyAddSpecies used
# datapath <- "/home/jvb740/AssignTaxonomy/"
# load(file.path(datapath, "FilteredSeqtab.RData"))
# 
# functpath <- "/home/jvb740/Dada_Pipel"
# 
# source(file.path(functpath, "Dada_TaxonomyFunctions.R"))
# 
# seqtab <- f_seqtab
# 
# if(!exists("seqtab")){
#         stop("no seqtab has been loaded")
# }
# 
# assignTaxonomyaddSpecies(seqtab, 
#                          minBoot = 80,
#                          allowMultiple = 3,
#                          PathToRefs = "/home/jvb740/AssignTaxonomy/",
#                          RefDataBase = "silva_nr_v123_train_set.fa.gz",
#                          SpeciesDB = "silva_species_assignment_v123.fa.gz",
#                          PathToSave = "/home/jvb740/AssignTaxonomy/Silva/")
# 
# assignTaxonomyaddSpecies(seqtab, 
#                          minBoot = 80,
#                          allowMultiple = 3,
#                          PathToRefs = "/home/jvb740/AssignTaxonomy/",
#                          RefDataBase = "gg_13_8_train_set_97.fa.gz",
#                          SpeciesDB = "silva_species_assignment_v123.fa.gz",
#                          PathToSave = "/home/jvb740/AssignTaxonomy/GG/")
# 
# assignTaxonomyaddSpecies(seqtab, 
#                          minBoot = 80,
#                          allowMultiple = 3,
#                          PathToRefs = "/home/jvb740/AssignTaxonomy/",
#                          RefDataBase = "rdp_train_set_14.fa.gz",
#                          SpeciesDB = "rdp_species_assignment_14.fa.gz",
#                          PathToSave = "/home/jvb740/AssignTaxonomy/rdp/")

```

# Load seqtab and taxonomy data and source functions

```{r, message = FALSE, warning = FALSE}
functionpath <- "/Users/jvb740/MarieCurie_Work/BackgroundKnowledge/16S_Learning/Dada_Pipel"

source(file.path(functionpath, "Dada_TaxonomyFunctions.R"))
source(file.path(functionpath, "Dada_PlotFunctions.R"))


datapath <- "/Users/jvb740/MarieCurie_Work/BackgroundKnowledge/16S_Learning/AssignTaxonomy/"
load(file.path(datapath, "Silva/Taxonomy.RData"))
taxa.species.silva <- taxa.species
load(file.path(datapath, "GG/Taxonomy.RData"))
taxa.species.gg <- taxa.species
## NB: the gg alone resulted already in an all NA species column that I remove here
taxa.species.gg <- taxa.species.gg[,-7]

load(file.path(datapath, "rdp/Taxonomy.RData"))
taxa.species.rdp <- taxa.species


load(file.path(datapath, "FilteredSeqtab.RData"))
seqtab <- f_seqtab
min(colSums(seqtab > 0)/nrow(seqtab)) # all taxa in seqtab_f were present in at least 20% of the samples
```


# Compare the Assignments

```{r, message = FALSE, warning = FALSE}
# Number and percentage of assigned taxonomic levels for all amplicons
NAPerC.silva <- NAPerCForTaxLevel(taxa.species.silva)

knitr::kable(NAPerC.silva) # still only 26% of amplicons assigned 

NAPerC.gg <- NAPerCForTaxLevel(taxa.species.gg)

knitr::kable(NAPerC.gg) # still only 29% of amplicons assigned

NAPerC.rdp <- NAPerCForTaxLevel(taxa.species.rdp)

knitr::kable(NAPerC.rdp)

```


- so the assignment numbers are very alike, question is if the assignemnts are the same