---
title: "alpha diversity measures"
author: "Thorsten Brach"
date: "25/7/2017"
output: html_document
---

# You start with a phyloseq ps object

- Load here ps object from ManiAging pretest or DanFund

```{r}
seqtab <- as(otu_table(ps), "matrix")
```

# Some phyloseq basics

```{r}
# ntaxa(ps)
# nsamples(ps)
# sample_names(ps)
# sample_variables(ps)
# get_variable(ps, sample_variables(ps)[2])
# sample_data(ps)$Group
# rank_names(ps)
# colnames(tax_table(ps))
# get_taxa_unique(ps, "Phylum")
# sample_sums(ps)
# taxa_sums(ps) # abundance of the taxa over all samples (sum)
# get_taxa(ps, i = sample_names(ps)[1]) # so i = 1 refers to a sample
# get_sample(ps, taxa_names(ps)[1])

```

# Alpha Diversity

## use estimate_richness to calculate Richness, Observed, and Shannon

```{r, message = FALSE, warning = FALSE}
AlphaDiv <- phyloseq::estimate_richness(ps, measures = c("Shannon", "Chao1", "Observed"))
rownames(AlphaDiv) <- sample_names(ps)
# I prefer Richness over Observed
colnames(AlphaDiv)[colnames(AlphaDiv) == "Observed"] <- "Richness"
```

- NB: estimate_richness warns if no singletons are in the data, i.e. if !any(otu_table(ps) == 1) = TRUE (= !any(seqtab.nochim == 1))
    - NB: dada2 seqtabs tend to be without singletons, most likely because low abundance SVs are considered errors from more abundant SVs


### NB: estimate_richness uses a functions from the vegan package

- specifically vegan::estimateR (e.g. for Chao1)
- and vegan::diversity (for Shannon and simpson)
- just type estimate_richness and check

```{r, message = FALSE, warning = FALSE}
AlphaMeasures <- vegan::estimateR(seqtab)
AlphaMeasures <- t(data.frame(AlphaMeasures))
```


## Calculate alpha diversity measures self

### Richness

```{r, message = FALSE, warning = FALSE}
Richness <- rowSums(seqtab != 0)

AlphaDiv$Richness_self <- Richness
head(AlphaDiv)
```

###  Chao1:

- Chao1 makes use of singletons and doubletons see vignette of vegan package and maybe <http://www.coloss.org/beebook/I/gut-symbionts/2/2/4>
- without singletons it is the same as Richness
- se.chao1 is the standard error of the chao1

```{r, message = FALSE, warning = FALSE}
# how to calculate chao1 yourself with singletons
seqtab2 <- seqtab
seqtab2[seqtab2 > 2 & seqtab2 < 10] <- 1 # introduce singletons
Estimator <- ((rowSums(seqtab2 == 1))*((rowSums(seqtab2 == 1))-1))/(2*((rowSums(seqtab2 == 2))+1))
Chao1 <- Richness+Estimator
## NB: the Variance (Standard Error) estimation is also given in the vignette but I was to lazy to do that here too

pss <- ps
otu_table(pss) <- otu_table(seqtab2, taxa_are_rows = FALSE)
AlphDivpss <- phyloseq::estimate_richness(pss, measures = c("Chao1"))
head(cbind(AlphDivpss, Chao1))
```


###  Shannon diversity 

```{r, message = FALSE, warning = FALSE}
# first use vegan::diversity function
shannon_vegan <- vegan::diversity(seqtab, index = "shannon")

# Self
# Basically just: $-sum(p_{i}*log(p_{i})$.
seqtab_RA <- seqtab/rowSums(seqtab)
seqtab_RA_logT <- seqtab_RA*(-1)*(log(seqtab_RA))
shannon_self <- rowSums(seqtab_RA_logT, na.rm = TRUE)
head(cbind(AlphaDiv, Shannon_vegan = shannon_vegan, Shannon_self = shannon_self))
```

### Simpson

```{r, message = FALSE, warning = FALSE}

simpson_vegan <- vegan::diversity(seqtab, index = "simpson")
seqtab_RA2 <- seqtab_RA*seqtab_RA
simpson_self <- 1-rowSums(seqtab_RA2, na.rm = TRUE)
identical(simpson_vegan, simpson_self) # TRUE
```
